Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2019-12-22 analysis/paper/paper.Rmd
seperate code blocks and working on PCA result

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  dpi = 300,
  fig.path = "../figures/"
)
```

@@

The elliptic Fourier coefficients of `r total_no` pottery from three phases were calculated. Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power. The mean pottery shape of each phase was visualized by calculating the mean of the standardized Fourier coefficients within each group. 
Thin-plate splines compares the average shapes of vessels from each period to visualize the outline deformations required to pass from an extreme of one morphospace to the other.

```{r EFA-analysis}
library(Momocs)
library(cowplot)
jpgs_imported_coo <- readr::read_rds(here::here("analysis/data/derived_data/jpgs_imported_coo.rds"))

@@

`Post-European` <- mshp$poe %T>%  coo_plot( border="blue")
`Pre-European` <-  mshp$pre  %T>% coo_plot(border="red", plot.new=FALSE)
`Chinese` <-       mshp$chi  %T>% coo_plot(border="green", plot.new=FALSE)
legend("topright", 
       lwd=1,
       col=c("red", "blue", "green"), 
       legend=c( "Pre-European", "Post-European", "Chinese"))

# Thin plate spline analysis (TPS)
# Thin plate splines (TPS) was used to compare average shapes and visualise the outline deformations required to pass from an extreme of the morphospace to the other.
# Deformation grids

tps_grid(`Pre-European`, `Post-European`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
         amp = 1, grid.size = 18, legend = T)
dev.off()

tps_grid(`Pre-European`, `Chinese`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
         amp = 1, grid.size = 18, legend = T)
dev.off()

# Isodeformation lines
png(here::here("analysis/figures/tps-iso.png"), res = 600, w = 1200, h = 1800)
de_pre_post_plot <- 
  tps_iso(`Pre-European`, `Post-European`,  shp.lwd = c(2, 2), shp.border = c("red", "blue"),
          amp = 0.5, iso.nb = 10000, iso.levels = 12, legend.text = c("Pre-contact", "Post-contact"),
          palette = col_summer, grid = F)
dev.off()

de_pre_chi_plot <- 
  tps_iso(`Pre-European`, `Chinese`,  shp.lwd = c(2, 2), shp.border = c("red", "blue"),
          amp = 0.5, iso.nb = 10000, iso.levels = 12, legend.text = c("Pre-contact", "Chinese"),
          palette = col_summer, grid = F)
dev.off()

# Doesn't work if the plot is not produced by ggplot?
# plot_two_de_by_period <- 
  #plot_grid(de_pre_post_plot, de_pre_chi_plot, labels = c('A', 'B'))

#knitr::include_graphics(here::here("analysis", "figures", "tps-pre-Euro-to-post-Euro.png"))
#knitr::include_graphics(here::here("analysis", "figures", "tps-pre-Euro-to-Chinese.png"))

```{r PCA-analysis}
# these are fast
op.p <- PCA(op)
# proportion of variances explained, PC1 and PC2 explain 79% of variation
pro_vari <-scree_plot(op.p)$data
vari_two <- round(pro_vari$cumsum[2],4)*100
vari_three <- round(pro_vari$cumsum[3],4)*100

# the pots groups by phase, and their shape distribution according to 
# the first two principle components of the EFA 
png(here::here("analysis/figures/pca-biplot.png"), res = 300, w = 1500, h = 1500)
plot(op.p, 2, 
     morpho = TRUE, 
     ellipses = TRUE,
     cex = 1, 
     cex.labelsgroups = 1.5)

dev.off()

# different visualization 
plot(PCA(op), 2, pos.shp="full")
plot(PCA(op), 2, pos.shp="range")
plot(PCA(op), 2, pos.shp="xy")
plot(PCA(op), 2, pos.shp="range_axes")

plot(PCA(op), 2, pos.shp="full_axes")

# the contributions of each PC to the shapes
pc_contrib <- PCcontrib(op.p, nax = 1:3) 
pc_contrib$gg +  
  geom_polygon(fill="slategrey", col="black") + 
  ggtitle("Shape variation along PC axes") +
  theme_minimal(base_size = 6)

@@

Outlines generated by EFA are examined by principal components analysis to explore the shape variations across three phases. The first two principal components (PCs) explain `r vari_two`% of variances, and the first three principal components (PCs) explain `r vari_three`% of variances. The PC1 relate to the height of the vessels from short to high, the PC2 reflects the neck and mouth constriction from wide to narrow, and the PC3 relates to the degree of flare of neck from curved to straight shape. The biplot of PCs presents the grouping by each occupation phase.

@@

manv <-   MANOVA_PW(op.p, "phase")
manv$summary

>>>>

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  dpi = 300,
  fig.path = "../figures/"
)

image_width_1_col_inch = 90/25.4 # image Single column	w = 90 mm 
image_width_2_col_inch = 190/25.4 # image Double column w = 190 mm 
```

@@

The elliptic Fourier coefficients of `r total_no` pottery from three phases were calculated. Reliable pottery outline was captured by 13 harmonics that gather 99 % of the total harmonic power. The average shape of vessels from each phase was visualized with the mean of the standardized Fourier coefficients within each phase group (Figure \@ref(fig:EFA-analysis)).

```{r EFA-analysis, fig.cap= "Mean pottery shapes of three phases"}
library(Momocs)
library(cowplot)
jpgs_imported_coo <- readr::read_rds(here::here("analysis/data/derived_data/jpgs_imported_coo.rds"))

@@

`Post-European` <- mshp$poe %T>% coo_plot( border="blue")
`Pre-European` <-  mshp$pre %T>% coo_plot(border="red", plot.new=FALSE)
`Chinese` <-       mshp$chi %T>% coo_plot(border="green", plot.new=FALSE)
legend("topright", 
       lwd=1,
       col=c("red", "blue", "green"), 
       legend=c( "Pre-European", "Post-European", "Chinese"))
```

```{r thin-plate-splines}
# Thin plate spline analysis (TPS)
# Thin plate splines (TPS) was used to compare average shapes and visualise the outline deformations required to pass from an extreme of the morphospace to the other.
# Deformation grids

tps_grid(`Pre-European`, `Post-European`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
           amp = 1, grid.size = 18, legend = T)

tps_grid(`Pre-European`, `Chinese`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
           amp = 1, grid.size = 18, legend = T)

# Isodeformation lines
png(here::here("analysis/figures/tps-iso.png"), res = 600, w = 1200, h = 1800)

# Doesn't work if the plot is not produced by ggplot?
# plot_two_de_by_period <- 
  #plot_grid(, labels = c('A', 'B'))

#knitr::include_graphics(here::here("analysis", "figures", "tps-pre-Euro-to-post-Euro.png"))
```

```{r PCA-table}
# these are fast
op.p <- PCA(op)
# proportion of variances explained, PC1 and PC2 explain 79% of variation
pro_vari <-scree_plot(op.p)$data
vari_two <- round(pro_vari$cumsum[2],4)*100
vari_three <- round(pro_vari$cumsum[3],4)*100
```

Outlines generated by EFA are examined by principal components analysis to explore the shape variations across three phases. The first two principal components (PCs) explain `r vari_two`% of variances, and the first three principal components (PCs) explain `r vari_three`% of variances. The PC1 relate to the height of the vessels from short to high, the PC2 reflects the neck and mouth constriction from wide to narrow, and the PC3 relates to the degree of flare of neck from curved to straight shape. The contribution of each principal component on the shape of vessels was listed in table (). The PC1 relative to the hight of the vessels accounts for nearly half of variances. Figure\@ref(fig:first-two-PCs-byplot) shows that a wider variation in pottery shape from the pre-European and post-European period compared to those from Chinese period.

```{r contribution-PCs}
# the contributions of each PC to the shapes
pc_contrib <- PCcontrib(op.p, nax = 1:3) 
pc_contrib$gg +  
  geom_polygon(fill="slategrey", col="black") + 
  ggtitle("Shape variation along PC axes") +
  theme_minimal(base_size = 6)

@@

```{r first-two-PCs-byplot, fig.cap= "Pottery shape distribution by each occupation phase according to the first two PCs"}
# the pots groups by phase, and their shape distribution according to 
# the first two principle components of the EFA 
png(here::here("analysis/figures/pca-biplot.png"), res = 300, w = 1500, h = 1500)
plot(op.p, 2, 
     morpho = TRUE, 
     ellipses = TRUE,
     cex = 1, 
     cex.labelsgroups = 1.5)
dev.off()

# PCA byplot
plot(PCA(op), 2, pos.shp="range")
```

@@

manv <-   MANOVA_PW(op.p, "phase")
# manv$summary
